# Mapping

## SOmap

## Supporting data

When constructing maps, we commonly want to show features like oceanographic fronts, ice extent, coastline, place names, and MPA boundaries. There are a few sources of such data:

- some layers are bundled into `SOmap`, see the `SOmap::SOmap_data` object
- `antanym` provides access to the SCAR Composite Gazetteer of place names
- the `quantarcticR` package provides access to [Quantarctica](http://quantarctica.npolar.no/) data layers.

### quantarcticR

Note, this package is still in development. Install if needed:

```{r}
remotes::install_github("SCAR-sandpit/quantarcticR")
```

Example usage:

```{r}
library(quantarcticR)
ds <- qa_datasets() ## all available layers
head(ds)

## more info about a particular layer
my_layer <- qa_dataset("Median sea ice extent 1981-2010")
my_layer

## fetch the actual data for that layer
layer_data <- qa_get(my_layer)

## plot it
plot(layer_data[layer_data$MONTH == "October", ])

## or add to a SOmap
SOmap(trim = -50, border_width = 0.5)
SOplot(layer_data[layer_data$MONTH == "October", ], col = "red")

```

### antanym

See the [overview](overview.html) article.
