---
title: "SCAR-EGABI Tools for Southern Ocean Spatial Analysis and Modelling"
author: Anton Van de Putte, Charlène Guillaumot, Grant Humphries, Huw Griffith, Ben Raymond
description: Course material
documentclass: book
site: bookdown::bookdown_site
---

```{r setup, echo = FALSE, cache = FALSE}
library(knitr)

knitr::opts_chunk$set(cache = TRUE)

## use this chunk to set up data directories and similar
#BOOKDOWN_DATA_DIR <- 
#library(quantarcticR)
## for vignette purposes, and if this is Ben's computer, use a persistent cache_dir
#if (grepl("ben_ray", qa_cache_dir())) qa_cache_dir("c:/data/Quantarctica3")
```

# About this

This document provides material for the [SCAR-EGABI Tools for Southern Ocean Spatial Analysis and Modelling Course](https://github.com/SCAR/EGABIcourse19/) to be held in Leuven, Belgium in September 2019.

Issues or suggestions about the content can be raised through the [issue tracker](https://github.com/SCAR/EGABIcourse19/issues) (GitHub account required).

## Technical stuff {-}

This material has been written using [bookdown](https://github.com/rstudio/bookdown) and [R](https://cran.r-project.org/).

To build the book locally, clone the [repo](https://github.com/SCAR/EGABIcourse19), open R and run the following lines:

```{r eval = FALSE}
setwd("the/path/to/your/copy/of/the/repo")
bookdown::render_book("index.Rmd")
```

And view it with:
```{r eval = FALSE}
browseURL("docs/index.html")
```

